---
title: "Factors that affect early childhood education"
author: "Win Suvansinpan, Geoffrey Li, Roshan Ramkeesoon"
date: "Wednesday, March 13, 2019"
output: beamer_presentation
fontsize: 9pt
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
setwd('~/ecls/')
set.seed(0)
library(tidyverse)
library(haven)
library(sjmisc)
library(ggplot2)
library(dplyr)


qqn <- function (x, ...) {
  qqnorm(x, ...)
  qqline(x)
}
```



# Outline

- Dataset

- Experiment 1. TV watching

- Experiment 2. Bedtime

- Experiment 3. Age

# Dataset

- Early Childhoold Longitudinal Study

- Sampling and data collection methdology

- Availability

- Common response variable

# Early Childhoold Longitudinal Study
- conducted by the Institute for Education Sciences, part of U.S. federal government \footnote{U.S. Dept of Education, National Center for Education Statistics. EARLY CHILDHOOD LONGITUDINAL STUDY [UNITED STATES]: KINDERGARTEN CLASS OF 1998-1999, THIRD GRADE [Computer file]. ICPSR version. Washington, DC: U.S. Dept of Education, Institute of Education Sciences [producer], 2004. Ann Arbor, MI: Inter-university Consortium for Political and Social Research [distributor], 2004.}
- collected data on factors that affect educational outcomes by interviewing students, their teachers, and their parents/guardians
- tracking 3,341 variables ranging from student bedtime to the teacher’s level of interest in teaching, the study is one of the largest datasets of its kind
- We only use the 2002 dataset, which pertains to 15,000 students in spring 3rd grade from the Kindergarten class of 1998-1999

# Sampling and data collection methdology
- 15,000 students nationwide
- multi-stage stratified cluster sampling
    - schools were randomly selected across U.S.A.
    - average of 23 kingergartners randomly selected from each selected school
- sample is diverse (socioeconomic status, race, private/public schools, etc)
- parents and teachers were interviewed over phone, students were interviewed in-person
- all interviews conducted by trained staff

# Availability
- ECLS dataset is available for download on the ECLS website  
- High level (>90%) of completness for most fields


# Common response variables

- math, reading, and science exams adminstered during student surveys
- "math t-score": raw scores in each subject were also provided as normalized to $N(\mu=50, sd=10)$
- normalization is useful so we can make comparisons in response across the three different subjects
- will examine average changes in child's normalized score associated with predictors






# Experiment 1. TV watching

- Background 

- Hypothesis

- Variables

- Method

- Results



# Background

- While television is a decades old technology, the interaction between time spent consuming media and child education outcomes is still of interest
- prior studies have shown that there is a statistically significant negative effect of early childhood television watching on educational outcomes at high school level
- prior studies indicate that time spent after classes on student activities and homework have significant positive influence on grades
- interested in matching prior study's association between total television watching with grades in our data
- interested in association of afterschool television watching on grades


# Hypotheses

- $H_{A0}$: there is no association between hours of television watching \textbf{total per week} on normalized student grades in any subject
- $H_{B0}$: there is no association between hours of television watching \textbf{after school} on normalized student grades in any subject

# Variables

- Control for parent income a numeric proxy of socioeconomic status which we think may have an effect on parent and child behavior

- Predictors are hours of tv watched in each period: 
    - weekdays before 8am
    - weekdays between 8am and 3pm
    - weekdays between 3pm and dinner
    - weekdays between dinner and bedtime
    - saturdays
    - sundays


# Method
- To test $H_{A0}$: Use linear regression to test for significance of hours of tv watched \textbf{total per week} on each subject score while controling for parent income
- To test $H_{B0}$: Use linear regression to test for significance of hours of tv watched \textbf{after school} on each subject score while controling for parent income


```{r, echo=F}
tv <- read.csv(file="data-processed/tv.csv", header=TRUE, sep=",")
```




# Television watched weekly
\small
```{r, echo=F, comment = ""}
tv.lr.math_sum <- lm(c5r2mtsc ~ I(p5tvbf8h+p5tvw83h+p5tvw3dh+p5tvafdh+p5tvsath+p5tvsunh)+w3income, data=tv)
tv.lr.read_sum <- lm(c5r2rtsc ~ I(p5tvbf8h+p5tvw83h+p5tvw3dh+p5tvafdh+p5tvsath+p5tvsunh)+w3income, data=tv)
tv.lr.scie_sum <- lm(c5stscor ~ I(p5tvbf8h+p5tvw83h+p5tvw3dh+p5tvafdh+p5tvsath+p5tvsunh)+w3income, data=tv)


tv_watched <- summary(tv.lr.math_sum)$coef[2,]
income <- summary(tv.lr.math_sum)$coef[3,]
print('MATH')
tv.lr.math.table <- rbind(tv_watched, income)
tv.lr.math.table


tv_watched <- summary(tv.lr.read_sum)$coef[2,]
income <- summary(tv.lr.read_sum)$coef[3,]
print('READING')
tv.lr.read.table <- rbind(tv_watched, income)
tv.lr.read.table


tv_watched <- summary(tv.lr.scie_sum)$coef[2,]
income <- summary(tv.lr.scie_sum)$coef[3,]
print('SCIENCE')
tv.lr.scie.table <- rbind(tv_watched, income)
tv.lr.scie.table

```
- All tv watching coefficients statistically significant, have negative association with test scores
- Income coefficient is statistically significant, positive association with grades
- Each additional hour of tv watched per week is associated with an average 0.05 to 0.07 drop in mean test scores for a given income
- We have reproduced the result that we reject the null hypothesis that there is no association between hours of television watched and grades

# Assess model assumptions
\small
```{r, echo = F}
par(mfrow=c(2,2))
plot(tv.lr.math_sum)
```

# Assess model assumptions
- slight deviance from normality in the residuals vs fitted plot
- qqplot shows normality of residuals
- scale-location plot shows constant variance of residuals across fitted values
- striation patterns in the residuals versus fitted plot might be explained by large jumps in data for parent's income
- similar results for comparisons with the other two tests (see report)
- seems like assumptions hold sufficiently for parameters of interest to be meaningful




# Television watched after school
\small
```{r, echo=F, comment = ""}
tv.lr.math_3 <- lm(c5r2mtsc ~ p5tvw3dh+w3income, data=tv)
tv.lr.read_3 <- lm(c5r2rtsc ~ p5tvw3dh+w3income, data=tv)
tv.lr.scie_3 <- lm(c5stscor ~ p5tvw3dh+w3income, data=tv)

math <- summary(tv.lr.math_3)$coef[2,]
reading <- summary(tv.lr.read_3)$coef[2,]
science <- summary(tv.lr.scie_3)$coef[2,]

tv_watched <- summary(tv.lr.math_3)$coef[2,]
income <- summary(tv.lr.math_3)$coef[3,]
print('MATH')
tv.lr.math.table <- rbind(tv_watched, income)
tv.lr.math.table


tv_watched <- summary(tv.lr.read_3)$coef[2,]
income <- summary(tv.lr.read_3)$coef[3,]
print('READING')
tv.lr.read.table <- rbind(tv_watched, income)
tv.lr.read.table


tv_watched <- summary(tv.lr.scie_3)$coef[2,]
income <- summary(tv.lr.scie_3)$coef[3,]
print('SCIENCE')
tv.lr.scie.table <- rbind(tv_watched, income)
tv.lr.scie.table
```
- All the coefficients are statistically significant and negative
- Income is significantly significant and positive association with grades
- Each additional hour of tv watched after school is associated with an average 0.23 to 0.37 drop in mean test scores for a given income
- We reject the null hypothesis that after school television watching does not have an association with grades


# Assess model assumptions - math scores
\small
```{r, echo = F}
par(mfrow=c(2,2))
plot(tv.lr.math_3)
```

# Assess model assumptions
- slight deviance from normality in the residuals vs fitted plot
- qqplot shows normality of residuals
- scale-location plot shows constant variance of residuals across fitted values
- striation patterns in the residuals versus fitted plot might be explained by large jumps in data for parent's income
- similar results to reading and science (see report)
- seems like assumptions hold sufficiently for parameters of interest to be meaningful



# Compare coefficients from both experiments
- Coefficients found from the two linear regression models mentioned on prevous slides:
```{r, echo=F, comment = ""}

x <- data.frame("Subject"=c("Math", "Reading", "Science"), 
                        "Weekly Hours"=c(-5.36e-02, -6.60e-02, -5.81e-02), 
                        "AfterSchool Hours"=c(-.23, -0.38, -.28))
x
```
- The coefficients are higher than for total television watching indicating the isolated effect of each “after school” tv watching hour associated with worse outcomes than each hour weekly tv watching




